### 2019-09-26 ch9_discussion/content.tex
update

%dus was ik aan het nadenken over de stand van zaken na dynbenchmark ─ oké, ik heb zin om een dynbenchmark2 te doen moet, alles meer en beter
%maar hoe hebben we een impact gehad? als je kijkt naar course material over trajectory inference zou ik zeggen, veel ─ we zijn een gemakkelijke bron van guidelines
%maar als ik kijk naar de manuscripten van nieuwe TI methoden zou ik zeggen dat we nog niet zijn waar we wouden geraken

\section{Self-assessment in trajectory inference}
Many trajectory inference (TI) tools lack quantitative assessment of the accuracy of the method. Instead, they rely on anecdotal evidence to demonstrate their added value. A brief review of 75 TI articles reveals that only about 37\% contain a self-assessment (Figure \ref{fig:benchmarks_over_time}). Peer-reviewed TI articles fared even worse, self-assessing in only 34\% of cases (n=55), whereas articles first published as a pre-print self-assess in 43\% of cases (n=39). Only three TI articles feature a comparison of at least 5 methods using 5 datasets or more \cite{sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,parra_reconstructingcomplexlineage_2018}.

Self-assessments are universally biased in favour of the authors, intentionally or otherwise, but this does not mean they should be skipped out on. Norel et al. provide a valuable discussion on overcoming overly optimistic self-assessments \cite{norel_selfassessmenttrapcan_2011}.
% TODO: improve working "this does not mean they should be skipped out on"

In this perspective, we hypothesise that low self-assessment rates are primarily caused by a lack of a standardised problem definition, benchmarking datasets, and metrics. 
We elaborate on these reasons and viable solutions to easy benchmarking trajectory inference.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/self_assessment.pdf} 
	\caption{\textbf{The number of manuscripts featuring trajectory inference tools over time.} Areas coloured in darker shades of blue represent manuscripts containing self-assessment of the tool, highlighting \textbf{A}: the number of datasets used and \textbf{B}: the number of methods used (including the method of the authors).}
	\label{fig:benchmarks_over_time}
\end{figure}

@@

@@

Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets (n=7). 
By now, already hundreds of suitable real datasets exist (Figure~\ref{fig:datasets}). Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes.

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/datasets.pdf} 
	\caption{\textbf{An overview collection of TI benchmarking datasets in function of their publication date and the topology of the trajectory.} These datasets are readily available on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.}
	\label{fig:datasets}
\end{figure}

\subsection{Multiple metrics}

@@

%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
%
%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.

\section{dyngen discussion}
% felt that upon the development of a new technologies, good quality control practices from
% other technologies were not being carried over because the data required to evaluate computational
% tools does not exist yet and is too expensive to develop.

\section{dynbenchmark discussion}
\begin{itemize}
	\item Already outdated when the manuscript was published online
	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
	\item Include RNA velocity as inputs
\end{itemize}

\section{SCORPIUS discussion}
\begin{itemize}
	\item Extension to inferring non-linear trajectories, i.e. with principal graphs
\end{itemize}

\section{bred discussion}

%\section{incgraph discussion}

%\section{dyno discussion}

>>>>

\mycomment{Note: This text was written very quickly and is only meant as a temporary coathanger.}

This chapter reflects on the impact of this work on the field, referring back to the research objectives that were defined in Chapter \ref{chap:introduction}.

dyngen is a succes, as we created a simulator of single cell data that is already used to evaluate trajectory\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019} and single cell network inference\cite{pratapa_benchmarkingalgorithmsgene_2019}.

dynbenchmark was meant to guide users to better practices for applying trajectory inference methods, and guide developers to adopt better practices for developing accurate and robust TI methods. The first part we succeeded in, as the guidelines are commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}. However, we do not observe an increase in developers of TI methods performing quantitative benchmarks. We will discuss this in more detail in section~\ref{sec:selfassessment}.

dynbenchmark was a benchmark of TI methods developed over the course of 4 years involving writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, developing novel metrics for comparing ground truth and predicted trajectories, writing a simulator for synthetic single cell data. Along the way, we have learned a lot about how to benchmark computational methods. We share our vision and guidelines on benchmarking computational methods in Chapter~\ref{chap:guidelines}.

Something about SCORPIUS.

Something about bred.

Something about dyno.

Something about gng?

\section{Self-assessment in trajectory inference} \label{sec:selfassessment}
Many articles introducing novel trajectory inference (TI) tools lack quantitative assessment of the accuracy of the method. Instead, they rely on anecdotal evidence to demonstrate their added value. A brief review of 75 articles reveals that only about 37\% contain a self-assessment (Figure \ref{fig:benchmarks_over_time}A,B). Peer-reviewed articles fared even worse, self-assessing in only 34\% of cases (n=55), whereas articles first published as a pre-print self-assess in 43\% of cases (n=39).

The number of datasets used and methods compared against is also below expectations (Figure \ref{fig:benchmarks_over_time}C,D). Only three TI articles feature a comparison of at least 5 methods using 5 datasets or more \cite{sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,parra_reconstructingcomplexlineage_2018}.

While self-assessments are universally biased in favour of the authors\cite{norel_selfassessmenttrapcan_2011} (intentially or not), it is dangerous and unusual to publish a computational tool without quantitatively demonstrating its performance compared to state-of-the-art methods. Indeed, a recent comparison of 45 TI methods demonstrated that most methods perform worse than a few baseline methods constructed by combining simple off-the-shelf algorithms such as PCA, $k$-means and MST\cite{saelens_comparisonsinglecelltrajectory_2019}.

In this perspective, we hypothesise that low self-assessment rates are primarily caused by a lack of a standardised problem definition, readily available benchmarking datasets, and suitable metrics. 
We elaborate on these causal reasons, and provide viable solutions for performing TI benchmarks more easily.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/self_assessment.pdf} 
	\caption{
		\textbf{Less than half of all TI articles perform quantitative self-assessment.} 
		\textbf{A:} Since 2016, the number of TI articles has been increasing rapidly. Note that TI methods with both a pre-print and a peer-reviewed article only count once in the overall tally.
		\textbf{B:} Less than 50\% of articles feature a self-assessment. Peer-reviewed articles self-assess only in 34\% of cases.
		\textbf{C:} The number of datasets used in each benchmark is low.
		\textbf{D:} The number of methods (inclusing itself) evaluated is low.
	}
	\label{fig:benchmarks_over_time}
\end{figure}

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/datasets.pdf} 
	\caption{\textbf{An overview collection of real TI benchmarking datasets in function of their publication date and the topology of the trajectory.} These datasets are readily available on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.}
	\label{fig:datasets}
\end{figure}

@@

Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets (n=7). 
By now, already hundreds of suitable real datasets are available from GEO and ArrayExpress (Figure~\ref{fig:datasets}). Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes.

@@

\subsection{Multiple metrics}
The largest problem, by far, seems to be a lack of standardised metrics to evaluate TI methods. None of the benchmarks use a metric directly aimed at comparing the topology or pairwise orderings of two trajectories.

Instead, most benchmarks (n=26) employ an ordering metric (e.g. pearson correlation) the pseudotime\footnote{The pseudotime value of is calculated by computing its distance from a pre-defined or inferred start cell.} of the predicted trajectory to ground truth information such as the cell type or time of sampling. Several benchmarks (n=5) use a clustering metric (e.g. ARI) to compare a cell's transition assignment or milestone assignment to the ground truth cell type. In four cases, multiple executions were compared to evaluate the robustness of the predictions.
In one particular instance, the pseudotime correlation metric was adapted to be suitable for comparing rooted trees\cite{street_slingshotcelllineage_2018}. In another, an internal measure is used to quantify the smoothness of gene expression along the pseudotime \cite{darocha_reconstructioncomplexsinglecell_2018}.

While each of these metrics provide some insight the correctness of cell orderings or cell clusterings, they do not evaluate the correctness of the prediction's topology. Using only a pseudotime-based metric to evaluate non-linear trajectories is particularly risky, as it

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.5\linewidth]{fig/metrics.pdf} 
	\caption{}
	\label{fig:metrics}
\end{figure}

@@

%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
%
%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality. 
%
%\section{dyngen discussion}
%% felt that upon the development of a new technologies, good quality control practices from
%% other technologies were not being carried over because the data required to evaluate computational
%% tools does not exist yet and is too expensive to develop.
%
%\section{dynbenchmark discussion}
%\begin{itemize}
%	\item Already outdated when the manuscript was published online
%	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
%	\item Include RNA velocity as inputs
%\end{itemize}
%
%\section{SCORPIUS discussion}
%\begin{itemize}
%	\item Extension to inferring non-linear trajectories, i.e. with principal graphs
%\end{itemize}
%
%\section{bred discussion}
%
%%\section{incgraph discussion}
%
%%\section{dyno discussion}
